---
title: "L1_Workshop"
author: "Patrick Kelly"
date: "Saturday, January 10, 2015"
output: html_document
---

Objective
---

Learn how to fit the classical linear model in the framework of generalized linear model `glm()`.
Compare the outputs of `lm()` and `glm()` and understand the differences between them.

```{r}
Linear.Model.Data <- read.csv("C:/Users/Patrick/SkyDrive/Documents/Education/UChicago/Linear_NonLinear/Week1ProjectData.csv",header=TRUE,sep=",")
Linear.Model.Data[1:10,]

Linear.Model.Data.Frame<-as.data.frame(Linear.Model.Data)
```

Fit Linear model Using lm() and glm()
---

Estimate the model using `lm()` and `glm()`

```{r}
Linear.Model.Data.lm<-lm(Output~Input,data=Linear.Model.Data.Frame)

Linear.Model.Data.glm<-glm(Output~Input,family=gaussian(link="identity"),data=Linear.Model.Data.Frame)

#Check and compare the names of the objects elements in both fits.
names(Linear.Model.Data.lm)

names(Linear.Model.Data.glm)

#Compare the summaries of outputs.
summary(Linear.Model.Data.lm)
summary(Linear.Model.Data.glm)
```

Compare the Model Objects.
---

Compare the following fields in the model objects.
Calculate the outputs of `glm()` using the fields of `lm()`

1. coefficients
2. residuals
3. fitted.values
4. linear.predictors
5. deviance
  1. Use deviance()
  2. Calculate deviance manually based on the definition given in the lecture
6. Akaike Information Criterion aic
  1. Obtain it by using AIC()
  2. Calculate it manually using the definition given in the lecture.
7. y
8. null.deviance
9. dispersion
10. The saturated model and the null model: log-likelihood and AIC

```{r}
#1
Linear.Model.Data.lm$coefficients

Linear.Model.Data.glm$coefficients

#2
Linear.Model.Data.lm$residuals[1:10]

Linear.Model.Data.glm$residuals[1:10]

#3
Linear.Model.Data.lm$fitted.values[1:10]

Linear.Model.Data.glm$fitted.values[1:10]

#4
Linear.Model.Data.lm$linear.predictors
  
Linear.Model.Data.glm$linear.predictors[1:10]

#5
  #1
  deviance(Linear.Model.Data.lm)

  deviance(Linear.Model.Data.glm)

  #2
  sum(Linear.Model.Data.lm$residuals^2)

  sum(Linear.Model.Data.glm$residuals^2)
  Linear.Model.Data.glm$deviance

#6
  #1
  AIC(Linear.Model.Data.lm)

  AIC(Linear.Model.Data.glm)

  #2
  (-2*deviance(Linear.Model.Data.lm) + 2*)
```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```



```{r}

```
