---
title: "L2_Workshop"
author: "Patrick Kelly"
date: "Saturday, January 17, 2015"
output: html_document
---

Objective 1
---

```{r}
#read in the data
sample.vector <- read.csv("C:/Users/Patrick/SkyDrive/Documents/Education/UChicago/Linear_NonLinear/sample_for_optimization.csv",header=TRUE,sep=",")

sample.vector <- sample.vector[,1]

#create fn for the log-likelihood
Negative.LL.Normal <- function(param.vector,sample){
  
  n <- length(sample)
  
  term1 <- (n*(0.5))*log(2*pi*param.vector[2])
  term2 <- (1/(2*param.vector[2]))*(sum((sample-param.vector[1])^2))
  
  return(term1 + term2)
  
}
```

Optimize the negative log-likelihood function using `optim()`

```{r}
Optimized.Negative.Log.Likelihood.optim<-optim(c(7,4),
                                               Negative.LL.Normal,
                                               sample=sample.vector,
                                               method="L-BFGS-B",
                                               hessian=TRUE,
                                               lower=c(-Inf,0),
                                               control=list(trace=1))

names(Optimized.Negative.Log.Likelihood.optim)

```

Observe the variables in object `Optimized.Negative.Log.Likelihood.optim`

```{r}
Optimized.Negative.Log.Likelihood.optim$par
Optimized.Negative.Log.Likelihood.optim$value
Optimized.Negative.Log.Likelihood.optim$counts
Optimized.Negative.Log.Likelihood.optim$convergence
```

**convergence**
An integer code. 0 indicates successful completion (which is always the case for "SANN" and "Brent"). Possible error codes are

1 indicates that the iteration limit maxit had been reached.
10 indicates degeneracy of the Nelder–Mead simplex.
51 indicates a warning from the "L-BFGS-B" method; see component message for further details.
52 indicates an error from the "L-BFGS-B" method; see component message for further details.

```{r}
Optimized.Negative.Log.Likelihood.optim$message
Optimized.Negative.Log.Likelihood.optim$hessian
```

Make a filled contour plot of the negative log-likelihood function.

```{r}
#Plot the objective function
Data.To.Plot.x<-seq(from=5,to=15,length.out=50)
Data.To.Plot.y<-seq(from=15,to=35,length.out=50)


Term.1<-length(sample.vector)/2*outer(log(2*pi*Data.To.Plot.y),rep(1,length(Data.To.Plot.y)))
Term.2<-outer(rep(1,length(Data.To.Plot.x)),unlist(lapply(Data.To.Plot.x,FUN=function(vector.element,Data.Vector) sum((Data.Vector-vector.element)^2),sample.vector)))*outer(1/2/Data.To.Plot.y,rep(1,length(Data.To.Plot.y)))
  
Negative.Log.Likelihood.Data<-Term.1+Term.2
filled.contour(Data.To.Plot.x,Data.To.Plot.y,Negative.Log.Likelihood.Data,color.palette=rainbow,nlevels=20)
```



```{r}

```



```{r}

```
