---
title: "LNL_HW_week5"
author: "Patrick Kelly"
date: "Saturday, February 14, 2015"
output: pdf_document
---

**This assignment helps understanding Poisson regression and Negative Binomial regression.**
---

Recent outbreak of measles caused debated about necessity of vaccination.

The data file for this project contains CDC data on immunization coverage for MMR for all U.S. states and the national immunization rate.

The data for outbreaks in each state were simulated using the assumption that after vaccination 90% of vaccinated children become immuned.

Another assumption was made that 100% of not immunized people get infected when exposed to the virus.

The data for this project are in the file `MeaslesImmunizationCoverageAndOutbreaks.csv`

Read the data.

```{r}
measles.data<-read.csv(file="C:/Users/Patrick/Documents/R/UChicago/Linear_NonLinear/MeaslesImmunizationCoverageAndOutbreaks.csv")

measles.data<-as.data.frame(measles.data)
```

Fit Poisson Regression
---

```{r}
measles.poisson.model<-glm(Outbreaks~Coverage,family=poisson,data=measles.data)
summary(measles.poisson.model)

names(measles.poisson.model)

measles.poisson.model$linear.predictors
```

Compare the Outbreaks data with the fitted values.

```{r}
library(ggplot2)

compare.results <- data.frame(rownum = c(1:length(measles.data$Outbreaks)), actual = measles.data$Outbreaks, fitted = measles.poisson.model$fitted.values)
compare.results

ggplot(compare.results, aes(rownum)) + geom_line(aes(y = actual, colour = "actual")) + geom_line(aes(y = fitted, colour = "fitted"))

#compare.results.long <- melt(compare.results, id="rownum")
#ggplot(data=compare.results.long,
#       aes(x=rownum, y=value, colour=variable)) +
#       geom_line()
```

Check that the link in our case is logarithmic.

```{r}
log(measles.poisson.model$fitted.values)
```

Interpretation of the model.

What if coverage changes by 1%?

Then the change of intensity measured in percentage points is

```{r}
(1-exp(measles.poisson.model$coef[2]*.01))*100
```

Fit Negative Binomial regression
---

The standard `glm()` function does not have functionality to fit negative binomial distribution. But the package MASS does have it.

```{r}
library(MASS)
```

Learn how to use the package to fit negative binomial regression.
Fit the model.
Interpret the results. Compare the fit with the fitted Poisson regression model.

First, fit the negative binomial where the disperions (k) = 1

```{r}
measles.negbinom.model <- glm(Outbreaks~Coverage, negative.binomial(1), data=measles.data)
measles.negbinom.model
```

Now, fit the negative binomial but let the model find the optimal dispersion

```{r}
measles.negbinom.model2 <- glm.nb(Outbreaks~Coverage, data=measles.data)
measles.negbinom.model2
summary(measles.negbinom.model2)

#compare AIC values of both binomial models --- the second model, using glm.nb, is better 
AIC(measles.negbinom.model,measles.negbinom.model2)

#now compare AIC values btwn the Poisson model and neg. binom. model
AIC(measles.poisson.model,measles.negbinom.model2)
```

**Interpret the results. Compare the fit with the fitted Poisson regression model**

Initially when observing the output from the Poisson model, we notice that the null deviance is not greater than the null degrees of freedom. This is a first quick check that suggests that there is not overdispersion. But we explore further through fitting the negative binomial model anyway. Through the use of `glm.nb` we notice that it does not converge (iteration limit reached) - this also suggests that there is not overdispersion in the distribution.

In further evaluating The negative binomial model we observe a significant negative correlation between the coverage and outbreaks (coeff = -.78), similar to the Poisson model. Both model fits are producing very similiar results but all other evidence suggests that the Poisson distribution / model is a better fit. This is further confirmed when comparing both models AIC values - again they are very simliar but the Poisson model is slightly lower (400.72 vs 402.72).

In conclusion, the Poisson model is the most appropriate fit and the Coverage is a significant indep. variable w/ coeff -.783.