---
title: "Time_Series_HW_4"
author: "Patrick Kelly"
date: "Thursday, May 14, 2015"
output: pdf_document
---

Assignment Notes:

The daily data is from Illinois Dept of Transporation (IDOT) for I80E 1EXIT (the 2nd data column) - note each data point is an hourly count of the number of vehicles at a specific location on I80E.

Use the daily data for last 2 weeks of June 2013 to develop an ARIMA forecasting model.

Objective is to forecast the hourly counts for July 1.

The actual data file for July 1 is included for you to test your estimate.

**Part 1**

Use ARIMA(p,d,q) model to forecast. Find the model returned by R auto.arima(). Change the values of p and q and determine the best model using AICc and BIC. Do AICc and BIC select the same model as the best model?

First, I manipulated the data in MS Excel to get it into a usable format.

```{r}
#import the data
raw_data <- read.csv("~/R/UChicago/Time_Series/I80_EAST_data.csv")
#there were some accidentally included empty rows at the end - clean those up
raw_data <- raw_data[1:384,]
#convert date column to dates
raw_data$Date <- as.character(raw_data$Date)
raw_data$Date <- as.Date(raw_data$Date, "%m/%d/%Y")
#create value of date and time merged together
raw_data$date_time <- as.POSIXct(paste(raw_data$Date,raw_data$Time), format="%Y-%m-%d %H:%M")

#convert data into time series format
data_ts <- ts(raw_data$I80E_1EXIT, frequency = 24)

#subsetting data --- splitting between June and July
june_ts <- window(data_ts, start = c(1,1), end = c(15,24))
july_ts <- window(data_ts, start = c(16,1), end = c(16,24))

#fit the ARIMA(p,d,q) model using auto.arima()
library("forecast")

(aa_fit <- auto.arima(june_ts, seasonal = FALSE))

#select model based on best AICc and BIC values
#rather than changing p and q manually, I am using the auto.arima() to optimize for these different parameters
#optimize for AICc
(aa_fit2 <- auto.arima(june_ts, seasonal = FALSE, ic="aicc"))
#optimize for BIC
(aa_fit3 <- auto.arima(june_ts, seasonal = FALSE, ic="bic"))

```

Do AICc and BIC select the same model as the best model?

No, as the output above indicates, when the non-seasonal arima model is optimized for AICc versus BIC different models are produces. The fit optimized for AICc produces ARIMA(2,0,3), while the fit optimized for BIC produces ARIMA(2,0,2).

**Part 2**

Use day of the week seasonal ARIMA(p,d,q)(P,Q,D)s model to forecast for July 1 (which is a Monday)

```{r}
library(tseries)
#day of week seasonal model
#create new ts object with a frequency that aligns with the day of the week
data_ts_2 <- ts(raw_data$I80E_1EXIT, frequency = (24*7))

#subsetting data --- splitting between June and July
june_ts_2 <- window(data_ts_2, start = c(1,1), end = c(3,24))
july_ts_2 <- window(data_ts_2, start = c(3,25), end = c(3,48))

#fit model
check.aa.fit <- auto.arima(june_ts_2, seasonal = TRUE)
(day.fit <- arima(june_ts, order=c(0,1,2), seasonal = list(order=c(0,1,0), period = (168) )))

#forecast
forecast.7.1<-forecast(day.fit,h=24)
#check forecase output
forecast.7.1$mean

```

**Part 3**

Use hour of the day seasonal ARIMA (p,d,q)(P,D,Q)s model to forecast for the hours 8:00, 9:00, 17:00 and 18:00 on July 1

```{r}
#hour of the day seasonal model
check.aa.fit <- auto.arima(june_ts, seasonal = TRUE)
(hour.fit <- arima(june_ts, order=c(2,0,1), seasonal = list(order=c(2,0,0), period = (24) )))

#forecast
forecast2.7.1<-forecast(hour.fit,h=24)
#check forecase output for 8:00, 9:00, 17:00 and 18:00
forecast2.7.1$mean[c(8,9,17,18)]
```

**Part 4**

For the July 1 8:00, 9:00, 17:00 and 18:00 forecasts, which model is better (part 2 or part 3) ?

```{r}
#determine residuals based on the difference between actual and projected

#day of week model
(f1<-as.vector(forecast.7.1$mean[c(8,9,17,18)]))
(a1<-as.vector(july_ts_2[c(8,9,17,18)]))
abs(f1-a1)
#total error
sum(abs(f1-a1))

#hour of day model
(f2<-as.vector(forecast2.7.1$mean[c(8,9,17,18)]))
(a2<-as.vector(july_ts[c(8,9,17,18)]))
abs(f2-a2)
#total error
sum(abs(f2-a2))
```

The part 2 (day of week) model is better based on checking the residuals, as displayed above.
