---
title: "Time_Series_HW_5"
author: "Patrick Kelly"
date: "Thursday, May 21, 2015"
output: pdf_document
---

**Assignment #5 - ARIMA Beer Sales**
---

Load data from TSA package (the package is written by textbook authors Jonathan Cryer and Kung-Sik Chan).

```{r, message=FALSE}
library("TSA")
library("forecast")
library("zoo")
library("hydroGOF")

data(beersales)
```

The data is the monthly beer sales in millions of barrels, 01/1975 - 12/1990.

**Part 1**
use ARIMA(p,d,q) model to forecast beer sales for all months of 1990.

```{r}
#limit data to 1990
(beer.2 <- window(beersales, 1975, c(1989,12)))

#fit the ARIMA(p,d,q)
(fit.beer <- auto.arima(beer.2, seasonal = FALSE))
```

1A - Use the h-period in forecast() to forecast each month of 1990

```{r}
#forecast all months of 1990
(pred.1990 <- forecast(fit.beer,h=12))
```

1B - Use the monthly data as a continuous time series. Forecast for 1990 Jan, Plug forecast into the time series to forecast for 1990 Feb. And so on and so forth.

```{r}
#initialize vectors to loop through each month
pred.loop.1990  <- vector(mode="list", length=13)
beer.loop <- vector(mode="list", length=13)
fit.beer.loop <- vector(mode="list", length=13)

beer.loop[[1]] <- beer.2
fit.beer.loop[[1]] <- fit.beer

#loop through each month, creating the forecast then merging the value onto the ts
for (i in 2:13 ) {
  
  #forecast on previous beer table
  (pred.loop.1990[[i]] <- forecast(fit.beer.loop[[i-1]],h=1)$mean)
  #merge on new forecasted value
  beer.loop[[i]] <- as.ts(c(as.zoo(beer.loop[[i-1]]),as.zoo(pred.loop.1990[[i]])))
  #fit a new model with the new value included
  (fit.beer.loop[[i]] <- auto.arima(beer.loop[[i]], seasonal = FALSE))
  
}

#check the final product
beer.loop[[13]]
```

1C - which of the two above approaches yield the better results in terms of Mean Squared Error 1990?

```{r}
#create ts of 1990 actuals
beer1990 <- window(beersales,1990)
  
#calc MSE for first forecast
mse(pred.1990$mean,beer1990)

#calc MSE for second forecast
mse(window(beer.loop[[13]],1990),beer1990)
```

Based on the MSE calculations above, the 2 forecasts product very similiar results but the first forecase (forecasting all 12 months at once) is marginally better.

**Part 2**
use month of the year seasonal ARIMA(p,d,q)(P,Q,D)s model to forecast beer sales for all the months of 1990.

```{r}
#fit the ARIMA(p,d,q)(P,Q,D)s
(fit.s.beer <- auto.arima(beer.2, seasonal = TRUE))

#forecast all months of 1990 with seasonality
(pred.s.1990 <- forecast(fit.s.beer,h=12))
```

**Part 3**
Which model (Part 1 or Part 2) is better to forecast beer sales for each month of 1990 (Jan, Feb, ..., Dec) ? 

```{r}
#calc MSE for part 1 forecast
mse(pred.1990$mean,beer1990)

#calc MSE for part 2 forecast
mse(pred.s.1990$mean,beer1990)
```

Based on the MSE calculations above, the part 2 forecast (the seasonal ARIMA) is considerably better. Using the seasonal ARIMA drastically reduces the MSE to ~0.57 (compared to the ~15.96 MSE from part 1)
